The Strucplot Framework:\\ Visualizing Multi-way Contingency Tables with \pkg{vcd}

### 2006-10-08 trunk/vcd/paper/Strucplot/JSS.Rnw
third bunch of changes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@583 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  David Meyer\\
  Department of Information Systems and Process Management\\
  E-mail: \email{David.Meyer@wu-wien.ac.at}\\
  URL: \url{http://wi.wu-wien.ac.at/~meyer/}\\

@@

    Basic functionality      &  $\times$  & $\times$  & $\times$ &$\times$ &$\times$ &   $\times$   & $\times$& $\times$\\
    Shape                    &            &           &          &         &$\times$ & $\times$  &  $\times$&\\
    Res.-based shadings  &  $\times$  &        & $\times$ & $\times$ & ($\times$) &     &($\times$)& ($\times$)\\
    Highlighting             &     &        &   &$\times$   &$\times$  & $\times$ & $\times$& $\times$\\
    Conditional views        &  $\times$  &        &    &$\times$ &          & $\times$ &  $\times$&\\
    Interaction              &     &        &   &  &$\times$  &   $\times$   & $\times$& $\times$\\
    Linking                  &     &        &   &  &$\times$  &       & $\times$& $\times$\\
    Extensible design        &     &        &  &$\times$ & &       &  &\\
    Language                 & \proglang{SAS} & \proglang{S} &
    \proglang{S} & \proglang{S} & \proglang{Java} & \proglang{XLisp} & \proglang{C++} & \proglang{Java}\\
    \hline
  \end{tabular}
  \caption{Comparison of current software environments.}
  \label{tab:compare}
\end{table}

@@

The strucplot framework is highly modularized: Figure~\ref{fig:struc}
shows the hierarchical relationship between the various components. 
On the lowest level, there are several groups of workhorse and 
parameter functions that directly or indirectly influence the final
appearance of the plot (see Table \ref{tab:grapcons} for an overview). 
These are examples of grapcon functions. They are created by generating functions
(\emph{grapcon generators}), allowing
flexible parameterization and extensibility (Figure~\ref{fig:struc}
only shows the generators). The generator names
follow the naming convention \code{\textit{group\_foo}()},  
where (\code{\textit{group}}) reflects the group the
generators belong to (strucplot core, labeling,
legend, shading, or spacing). The workhorse functions (created by 
\code{struc\_\textit{foo}()}, 
\code{labeling\_\textit{foo}()}, and \code{legend\_\textit{foo}()}) 
directly produce graphical output (i.e., ``add ink to the canvas''), whereas
the parameter functions
(created by \code{spacing\_\textit{foo}()} and \code{shading\_\textit{foo}()}) compute
graphical parameters used by the others. The grapcon functions returned by 
\code{struc\_\textit{foo}()} implement the core functionality, 
creating the tiles and their
content.  On the second level of the framework, a suitable combination
of the low-level grapcon functions (or, alternatively, corresponding generating functions)
is passed as ``hyperparameters'' to \codefun{strucplot}.
This central function 
sets up the graphical layout using grid viewports (see Figure~\ref{fig:layout}),
and coordinates the specified core, labeling, shading, and spacing functions to produce 
the plot. On the third level, we provide
several convenience functions such as \codefun{mosaic},
\codefun{sieve}, \codefun{assoc}, and \codefun{doubledecker} which 
interface \codefun{strucplot} through sensible parameter defaults
and support for model formulae. Finally, on the fourth
level, there are ``related'' \pkg{vcd} functions (such as \codefun{cotabplot}
and the \codefun{pairs} methods for table objects) 
arranging collections of plots of the strucplot 
framework into more complex displays (e.g., by means of panel functions).

@@

\noindent Alternatively, we can directly inspect the residuals. 
The Pearson residuals (standardized deviations of observed from expected values) 
are conveniently visualized using association
plots \citep{vcd:Cohen:1980}. In contrast to \codefun{assocplot} in
base \proglang{R}, \pkg{vcd}'s \codefun{assoc}
function scales to more than two variables (see Figure~\ref{fig:residuals}):

@@

For both mosaic plots and association plots, 
the splitting of the tiles can be controlled using the
\code{split\_vertical} argument. The default is to alternate splits starting
with a horizontal one (see Figure~\ref{fig:split}).

@@

\noindent \emph{Conditioning} on levels (i.e., choosing a table subset for fixed levels
of the conditioning variable(s)) is done using the \code{[[} operator. %]]
Here again, the sequence of conditioning levels is restricted by the
hierarchical structure of the \class{structable} object. In the
following examples, note that compared to subsetting, the first
dimension(s) have been dropped:

@@

\noindent The effect of using this is that
conditioning variables are permuted ahead of the the conditioned variables
in the table, and that \codefun{spacing\_conditional} is used as default to better
distinguish conditioning from conditioned dimensions.
This spacing uses equal space between tiles of conditioned variables,
and increasing space between tiles of conditioning variables.

Another set of high-level functions for visualizing conditional independence
models are the \codefun{pairs} methods for \class{table} and \class{structable} objects.
In contrast to \codefun{cotabplot} which conditions on variables,
the \codefun{pairs} methods create pairwise views of the table. 
The function produces, by default, a plot matrix 
having strucplot displays in the off-diagonal panels, and
the variable names (optionally, with univariate displays) in the diagonal cells. 
Figure~\ref{fig:pairs} shows a pairs display for the
\data{Titanic} data with univariate mosaics in the diagonal,
and mosaic plots visualizing the corresponding bivariate mosaics
in the upper and lower triangles.
Due to the inherent asymmetry of mosaic displays, the corresponding plots in the
upper and lower triangle differ depending on which variable is used first
for splitting---inspecting both views might help detecting patterns in a data set. 
Additionally, 
we are using a special spacing and shading normally used to `highlight' %'
the second variable in the first (as will be discussed in Section \ref{sec:spacing}): here,
the intention of the spacing 
is to emphasize the conditional distributions of the second variable,
given the first one, and the shading helps identifying the factor levels in
the second variable.

@@

\begin{enumerate}
\item recursively, create a tree of nested viewports
\item push the whole tree, i.e., add it to the system tree maintained
  by the \pkg{grid} engine
\item one by one, seek all cells of the strucplot display and add
  plot-specific content (rectangles for mosaics, bars for association
  plots, etc.)
\end{enumerate}

@@

\cite{vcd:Friendly:1994}
suggested a residual-based shading for the mosaic tiles that can also be applied
to the rectangles in association plots \citep{vcd:Meyer+Zeileis+Hornik:2003}.
Apart from \codefun{shading\_binary}, there are currently two basic
shadings available in \pkg{vcd}:
\codefun{shading\_hcl} and \codefun{shading\_hsv}, as well
as two derived functions: \codefun{shading\_Friendly} building upon
\codefun{shading\_hsv}, and \codefun{shading\_max} building upon \codefun{shading\_hcl}.
\codefun{shading\_hsv} and \codefun{shading\_hcl} provide
the same conceptual tools, but use different color spaces: the
Hue-Saturation-Value (HSV) and the Hue-Chroma-Luminance (HCL) scheme,
respectively. We will first expose the basic concept of these shading
functions using HSV space, and
then briefly explain the differences to HCL space \citep[a detailed
discussion can be found in][]{vcd:Zeileis+Meyer+Hornik:2005}. 
Color palettes in HCL space are 
preferable to palettes derived 
from HSV space from a perceptual point of view. Functions for creating
palettes (see, e.g., \codefun{diverge\_hcl}) are provided with the package.

@@

\noindent In Figure~\ref{fig:haireye}, the plot
in the middle depicts the default palette, 
and the bottom plot an alternative setting for Hue (\code{h}), Chroma (\code{s}), and
Luminance (\code{v}).

@@

\begin{figure}[htbp]
\begin{center}
<<haireyefig1,fig=TRUE,echo=FALSE,height=5>>=
mosaic(haireye, gp = shading_hsv, margin = c(bottom = 1), keep_aspect_ratio = FALSE)
@
<<haireyefig2,fig=TRUE,echo=FALSE,height=5>>=
mosaic(haireye, gp = shading_hcl, margin = c(bottom = 1), keep_aspect_ratio = FALSE)
@
<<haireyefig3,fig=TRUE,echo=FALSE,height=5>>=
mosaic(haireye, gp = shading_hcl, margin = c(bottom = 1), gp_args = list(h = c(130, 43), c = 100, l = c(90, 70)), keep_aspect_ratio = FALSE)
@
\caption{Three mosaic plots for the \data{HairEyeColor} data using, from top to bottom:
the default HSV color palette, the default HCL color palette, 
and a custom HCL color palette.}
\label{fig:haireye}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<label1fig,fig=TRUE,echo=FALSE,height=12,width=8>>=
pushViewport(viewport(layout = grid.layout(ncol = 2,nrow=3)))
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
mosaic(Titanic, newpage = FALSE, keep = TRUE, margin = c(right = 3))
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
mosaic(Titanic, labeling_args = list(clip = c(Survived = TRUE, Age = TRUE)), 
       newpage = FALSE, keep = TRUE, margin = c(left = 3))
popViewport()
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = TRUE, Age = 2)), 
       newpage = FALSE, keep = TRUE, margin = c(right = 3))
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 2))
mosaic(Titanic, labeling_args = list(rep = c(Survived = FALSE, Age = FALSE)), 
       newpage = FALSE, keep = TRUE, margin = c(left = 3))
popViewport()
pushViewport(viewport(layout.pos.col = 1:2, layout.pos.row = 3))
pushViewport(viewport(width = 0.55))
mosaic(Titanic, labeling_args = list(rot_labels = c(bottom = 90, right = 0), 
                  offset_varnames = c(right = 1), offset_labels = c(right = 0.3)), 
       margins = c(right = 4, bottom = 3), newpage = FALSE, keep = FALSE)
popViewport(3)
@
\caption{Labeling the Titanic data. Top left: default labeling, top
  right: clipping, middle left: abbreviating, middle right: inhibiting 
  the repetition of levels, bottom: rotating.}
\label{fig:labels1}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

\noindent The \var{tl\_foo} (``top left'') arguments are \code{TRUE} by default. 
Now, we will add boxes to the labels and additionally
enable clipping (see Figure~\ref{fig:labels2}, bottom left):

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<label2fig,fig=TRUE,echo=FALSE,height=8,width=8>>=
pushViewport(viewport(layout = grid.layout(ncol = 2, nrow = 2)))
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
mosaic(Titanic, labeling_args = list(pos_varnames = "left", pos_labels = "left", 
       just_labels = "left", rep = FALSE), newpage = FALSE, keep = TRUE)
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, 
                  abbreviate = c(Survived = 1, Age = 3)), newpage = FALSE, keep = TRUE)
popViewport()
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, 
       boxes = TRUE, clip = TRUE), newpage = FALSE, keep = TRUE)
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 2))
mosaic(Titanic, labeling_args = list(tl_labels = TRUE, boxes = TRUE,
       clip = c(Survived = FALSE, TRUE), 
       labbl_varnames = TRUE), margins = c(left = 4, right = 1, 3), 
       newpage = FALSE, keep = TRUE)
popViewport(2)
@
\caption{Labeling the Titanic data, advanced features. Top left: left aligning, top
  right: changes in the margins, bottom left: boxes and clipping, bottom right: 
  Variable names beneath levels, and clipping disabled for the survival variable.}
\label{fig:labels2}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<label3fig,fig=TRUE,echo=FALSE,height=8,width=8>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(ncol = 2, nrow = 2)))
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
mosaic(~ MaritalStatus + Gender, data = PreSex, labeling = labeling_cells, newpage = FALSE, keep = TRUE)
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
mosaic(~ PremaritalSex + ExtramaritalSex, data = PreSex, 
       labeling = labeling_cells(abbreviate_labels = TRUE, abbreviate_varnames = TRUE, 
         clip = FALSE), newpage = FALSE, keep = TRUE)
popViewport()
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
mosaic(~ PremaritalSex + ExtramaritalSex | MaritalStatus + Gender, data = PreSex, 
       labeling = labeling_conditional(abbreviate_varnames = TRUE, 
         abbreviate_labels = TRUE, clip = FALSE, gp_text = gpar(col = "red")), 
       newpage = FALSE, keep = TRUE)
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 2))
mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = 1, Age = 4)), 
       pop = FALSE, newpage = FALSE, keep = TRUE)
tab <- ifelse(Titanic < 6, NA, Titanic)
labeling_cells(text = tab, clip = FALSE)(Titanic)
@
\caption{Cell labeling. Top left: default labeling using the
  \data{PreSex} data, top right: abbreviated labels, bottom left:
  conditional labeling (labels abbreviated and in red for clarity),
  bottom right: user-supplied text (observed frequencies exceeding 5) 
  added to a mosaic display of the \data{Titanic} data.}
\label{fig:labels3}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<listfig,fig=TRUE,echo=FALSE>>=
<<list>>
@
\caption{Labels indicated below the plot.}
\label{fig:list}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<spacingfig,fig=TRUE,echo=FALSE,height=8,width=8>>=
pushViewport(viewport(layout = grid.layout(ncol = 2, nrow = 2)))
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
mosaic(art, spacing = spacing_equal(unit(2, "lines")), keep = TRUE, newpage = FALSE)
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
mosaic(art, spacing = spacing_dimequal(unit(c(2,0.5), "lines")), keep = TRUE, newpage = FALSE)
popViewport()
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
mosaic(art, spacing = spacing_increase(start = unit(0.5, "lines"), rate = 1.5), keep = TRUE, newpage = FALSE)
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 2))
mosaic(art, spacing = spacing_highlighting, keep = TRUE, newpage = FALSE)
popViewport(2)
@
\caption{Varying spacing for the Arthritis  data. Top left: equal
  spacing for all dimensions, top right: different spacings for
  individial dimensions, bottom left: increasing spacing, bottom right: 
  spacing used for highlighting.}
\label{fig:spacing}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

>>>>

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  David Meyer\\
  Department of Information Systems and Operations\\
  E-mail: \email{David.Meyer@wu-wien.ac.at}\\
  URL: \url{http://wi.wu-wien.ac.at/~meyer/}\\

@@

    Basic functionality      &  $\times$  & $\times$  & $\times$ &$\times$ &$\times$ &   $\times$   & $\times$& $\times$\\
    Shape                    &            &           &          &$\times$ && $\times$  &  $\times$&\\
    Res.-based shadings  &  $\times$  &        & $\times$ & $\times$ & ($\times$) &     &($\times$)& ($\times$)\\
    Highlighting             &     &        &   &$\times$   &$\times$  & $\times$ & $\times$& $\times$\\
    Conditional views        &  $\times$  &        &    &$\times$ &          & $\times$ &  $\times$&\\
    Interaction              &     &        &   &  &$\times$  &   $\times$   & $\times$& $\times$\\
    Linking                  &     &        &   &  &$\times$  & $\times$ & $\times$& $\times$\\
    Extensible design        &     &        &  &$\times$ & &       &  &\\
    Language                 & \proglang{SAS} & \proglang{S} &
    \proglang{R} & \proglang{R} & \proglang{R}/\proglang{Java} & \proglang{XLisp} & \proglang{C++} & \proglang{Java}\\
    \hline
  \end{tabular}
  \caption{Comparison of current software environments.}
  \label{tab:compare}
\end{table}

@@

The strucplot framework is highly modularized: Figure~\ref{fig:struc}
shows the hierarchical relationship between the various components. 
On the lowest level, there are several groups of workhorse and 
parameter functions that directly or indirectly influence the final
appearance of the plot (see Table \ref{tab:grapcons} for an overview). 
These are examples of grapcon functions. They are created by generating functions
(\emph{grapcon generators}), allowing
flexible parameterization and extensibility (Figure~\ref{fig:struc}
only shows the generators). The generator names
follow the naming convention \code{\textit{group\_foo}()},  
where \code{\textit{group}} reflects the group the
generators belong to (strucplot core, labeling,
legend, shading, or spacing). The workhorse functions (created by 
\code{struc\_\textit{foo}()}, 
\code{labeling\_\textit{foo}()}, and \code{legend\_\textit{foo}()}) 
directly produce graphical output (i.e., ``add ink to the canvas''), whereas
the parameter functions
(created by \code{spacing\_\textit{foo}()} and \code{shading\_\textit{foo}()}) compute
graphical parameters used by the others. The grapcon functions returned by 
\code{struc\_\textit{foo}()} implement the core functionality, 
creating the tiles and their
content.  On the second level of the framework, a suitable combination
of the low-level grapcon functions (or, alternatively, corresponding generating functions)
is passed as ``hyperparameters'' to \codefun{strucplot}.
This central function 
sets up the graphical layout using grid viewports (see Figure~\ref{fig:layout}),
and coordinates the specified core, labeling, shading, and spacing functions to produce 
the plot. On the third level, we provide
several convenience functions such as \codefun{mosaic},
\codefun{sieve}, \codefun{assoc}, and \codefun{doubledecker} which 
interface \codefun{strucplot} through sensible parameter defaults
and support for model formulae. Finally, on the fourth
level, there are ``related'' \pkg{vcd} functions (such as \codefun{cotabplot}
and the \codefun{pairs} methods for table objects) 
arranging collections of plots of the strucplot 
framework into more complex displays (e.g., by means of panel functions).

@@

\noindent where \code{spacing\_dimequal} is used to set the spacing
of the second and third dimension to zero. Alternatively, we can directly inspect the residuals. 
The Pearson residuals (standardized deviations of observed from expected values) 
are conveniently visualized using association
plots \citep{vcd:Cohen:1980}. In contrast to \codefun{assocplot} in
base \proglang{R}, \pkg{vcd}'s \codefun{assoc}
function scales to more than two variables (see Figure~\ref{fig:residuals}):

@@

For both mosaic plots and association plots, 
the splitting of the tiles can be controlled using the
\code{split\_vertical} argument. The default is to alternate splits starting
with a horizontal one (see Figure~\ref{fig:split}):

@@

\noindent \emph{Conditioning} on levels (i.e., choosing a table subset for fixed levels
of the conditioning variable(s)) is done using the \code{[[} operator. %]]
Here again, the sequence of conditioning levels is restricted by the
hierarchical structure of the \class{structable} object. In the
following examples, note that compared to subsetting, the first
dimension(s) are dropped:

@@

\noindent The effect of using this is that
conditioning variables are permuted ahead of the the conditioned variables
in the table, and that \codefun{spacing\_conditional} is used as default to better
distinguish conditioning from conditioned dimensions.
This spacing uses equal space between tiles of conditioned variables,
and increasing space between tiles of conditioning variables (See 
Section~\ref{sec:spacing}).

Another set of high-level functions for visualizing conditional independence
models are the \codefun{pairs} methods for \class{table} and \class{structable} objects.
In contrast to \codefun{cotabplot} which conditions on variables,
the \codefun{pairs} methods create pairwise views of the table. 
They produce, by default, a plot matrix 
having strucplot displays in the off-diagonal panels, and
the variable names (optionally, with univariate displays) in the diagonal cells. 
Figure~\ref{fig:pairs} shows a pairs display for the
\data{Titanic} data with univariate mosaics in the diagonal,
and mosaic plots visualizing the corresponding bivariate mosaics
in the upper and lower triangles.
Due to the inherent asymmetry of mosaic displays, the corresponding plots in the
upper and lower triangle differ depending on which variable is used first
for splitting---inspecting both views might help detecting patterns in a data set. 
Additionally, 
we are using a special spacing and shading normally used to `highlight' %'
the second variable in the first (as will be discussed in Section \ref{sec:spacing}): here,
the intention of the spacing 
is to emphasize the conditional distributions of the second variable,
given the first one, and the shading helps identifying the factor levels in
the second variable.

@@

\begin{enumerate}
\item recursively, split the table until the individual cells are reached
\item during the splits, add viewports to the plot
\item for the individual cells, add plot-specific content 
      (rectangles for mosaics, bars for association plots, etc.)
\end{enumerate}

@@

\cite{vcd:Friendly:1994}
suggested a residual-based shading for the mosaic tiles that can also be applied
to the rectangles in association plots \citep{vcd:Meyer+Zeileis+Hornik:2003}.
Apart from \codefun{shading\_binary}, there are currently two basic
shadings available in \pkg{vcd}:
\codefun{shading\_hcl} and \codefun{shading\_hsv}, as well
as two derived functions: \codefun{shading\_Friendly} building upon
\codefun{shading\_hsv}, and \codefun{shading\_max} building upon \codefun{shading\_hcl}.
\codefun{shading\_hsv} and \codefun{shading\_hcl} provide
the same conceptual tools, but use different color spaces: the
Hue-Saturation-Value (HSV) and the Hue-Chroma-Luminance (HCL) scheme,
respectively. We will first expose the basic concept of these shading
functions using HSV space, and
then briefly explain the differences to HCL space \citep[a detailed
discussion can be found in][]{vcd:Zeileis+Meyer+Hornik:2005}. 
Color palettes in HCL space are 
preferable to palettes derived 
from HSV space from a perceptual point of view. Functions for creating
palettes (see, e.g., \codefun{diverge\_hcl}) are provided with the
\pkg{vcd} package.

@@

\noindent In Figure~\ref{fig:haireye}, the plot
in the middle depicts the default palette, 
and the bottom plot an alternative setting for Hue (\code{h}), Chroma (\code{c}), and
Luminance (\code{l}).

@@

\begin{figure}[htbp]
\begin{center}
<<haireyefig1,fig=TRUE,echo=FALSE,height=3,width=5>>=
mosaic(haireye, gp = shading_hsv, margin = c(bottom = 1), keep_aspect_ratio = FALSE)
@
<<haireyefig2,fig=TRUE,echo=FALSE,height=3,width=5>>=
mosaic(haireye, gp = shading_hcl, margin = c(bottom = 1), keep_aspect_ratio = FALSE)
@
<<haireyefig3,fig=TRUE,echo=FALSE,height=3,width=5>>=
mosaic(haireye, gp = shading_hcl, margin = c(bottom = 1), gp_args = list(h = c(130, 43), c = 100, l = c(90, 70)), keep_aspect_ratio = FALSE)
@
\caption{Three mosaic plots for the \data{HairEyeColor} data using, from top to bottom:
the default HSV color palette, the default HCL color palette, 
and a custom HCL color palette.}
\label{fig:haireye}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<label1fig,fig=TRUE,echo=FALSE,height=10,width=8>>=
pushViewport(viewport(layout = grid.layout(ncol = 2,nrow=3)))
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
mosaic(Titanic, newpage = FALSE, keep = TRUE, margin = c(right = 3), 
       gp_labels = gpar(fontsize = 10))
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
mosaic(Titanic, labeling_args = list(clip = c(Survived = TRUE, Age = TRUE)), 
       newpage = FALSE, keep = TRUE, margin = c(left = 3), gp_labels = gpar(fontsize = 10))
popViewport()
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = TRUE, Age = 2)), 
       newpage = FALSE, keep = TRUE, margin = c(right = 3), gp_labels = gpar(fontsize = 10))
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 2))
mosaic(Titanic, labeling_args = list(rep = c(Survived = FALSE, Age = FALSE)), 
       newpage = FALSE, keep = TRUE, margin = c(left = 3), gp_labels = gpar(fontsize = 10))
popViewport()
pushViewport(viewport(layout.pos.col = 1:2, layout.pos.row = 3))
pushViewport(viewport(width = 0.5))
mosaic(Titanic, labeling_args = list(rot_labels = c(bottom = 90, right = 0), 
                  offset_varnames = c(right = 1), offset_labels = c(right = 0.3)), 
       margins = c(right = 4, bottom = 3), newpage = FALSE, keep = FALSE, gp_labels = gpar(fontsize = 10))
popViewport(3)
@
\caption{Labeling the Titanic data mosaic. Top left: default labeling, top
  right: clipping, middle left: abbreviating, middle right: inhibiting 
  the repetition of levels, bottom: rotating.}
\label{fig:labels1}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

\noindent The tl\_\var{foo} (``top left'') arguments are \code{TRUE} by default. 
Now, we will add boxes to the labels and additionally
enable clipping (see Figure~\ref{fig:labels2}, bottom left):

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<label2fig,fig=TRUE,echo=FALSE,height=8,width=8>>=
pushViewport(viewport(layout = grid.layout(ncol = 2, nrow = 2)))
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
mosaic(Titanic, labeling_args = list(pos_varnames = "left", pos_labels = "left", 
       just_labels = "left", rep = FALSE), newpage = FALSE, keep = TRUE,
       gp_labels = gpar(fontsize = 12))
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, 
                  abbreviate = c(Survived = 1, Age = 3)), newpage = FALSE, keep = TRUE,
       margins = c(left = 4, right = 1, 3), gp_labels = gpar(fontsize = 12))
popViewport()
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
mosaic(Titanic, labeling_args = list(tl_labels = FALSE, tl_varnames = TRUE, 
       boxes = TRUE, clip = TRUE), newpage = FALSE, keep = TRUE, 
       gp_labels = gpar(fontsize = 12))
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 2))
mosaic(Titanic, labeling_args = list(tl_labels = TRUE, boxes = TRUE,
       clip = c(Survived = FALSE, TRUE), 
       labbl_varnames = TRUE), margins = c(left = 4, right = 1, 3), 
       newpage = FALSE, keep = TRUE, gp_labels = gpar(fontsize = 12))
popViewport(2)
@
\caption{Labeling the Titanic data, advanced features. Top left: left aligning, top
  right: changes in the margins, bottom left: boxes and clipping, bottom right: 
  Variable names beneath levels, and clipping disabled for the survival variable.}
\label{fig:labels2}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<label3fig,fig=TRUE,echo=FALSE,height=8,width=8>>=
grid.newpage()
pushViewport(viewport(layout = grid.layout(ncol = 2, nrow = 2)))
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
mosaic(~ MaritalStatus + Gender, data = PreSex, labeling = labeling_cells, newpage = FALSE, keep = TRUE, gp_labels = gpar(fontsize = 10))
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
mosaic(~ PremaritalSex + ExtramaritalSex, data = PreSex, 
       labeling = labeling_cells(abbreviate_labels = TRUE, abbreviate_varnames = TRUE, 
         clip = FALSE), newpage = FALSE, keep = TRUE, gp_labels = gpar(fontsize = 10))
popViewport()
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
mosaic(~ PremaritalSex + ExtramaritalSex | MaritalStatus + Gender, data = PreSex, 
       labeling = labeling_conditional(abbreviate_varnames = TRUE, 
         abbreviate_labels = TRUE, clip = FALSE, gp_text = gpar(col = "red")), 
       newpage = FALSE, keep = TRUE, gp_labels = gpar(fontsize = 10))
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 2))
mosaic(Titanic, labeling_args = list(abbreviate = c(Survived = 1, Age = 3)), 
       pop = FALSE, newpage = FALSE, keep = TRUE, gp_labels = gpar(fontsize = 10))
tab <- ifelse(Titanic < 6, NA, Titanic)
labeling_cells(text = tab, clip = FALSE)(Titanic)
@
\caption{Cell labeling. Top left: default labeling using the
  \data{PreSex} data, top right: abbreviated labels, bottom left:
  conditional labeling (labels abbreviated and in red for clarity),
  bottom right: user-supplied text (observed frequencies exceeding 5) 
  added to a mosaic display of the \data{Titanic} data.}
\label{fig:labels3}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<listfig,fig=TRUE,echo=FALSE>>=
mosaic(Titanic, labeling = labeling_list, margins = c(bottom = 5), keep = TRUE)
@
\caption{Labels indicated below the plot.}
\label{fig:list}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}[p]
\begin{center}
<<spacingfig,fig=TRUE,echo=FALSE,height=8,width=8>>=
pushViewport(viewport(layout = grid.layout(ncol = 2, nrow = 2)))
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 1))
mosaic(art, spacing = spacing_equal(unit(2, "lines")), keep = TRUE, newpage = FALSE)
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 1))
mosaic(art, spacing = spacing_dimequal(unit(c(0.5, 2), "lines")), keep = TRUE, newpage = FALSE)
popViewport()
pushViewport(viewport(layout.pos.col = 1, layout.pos.row = 2))
mosaic(art, spacing = spacing_increase(start = unit(0.3, "lines"), rate = 2.5), keep = TRUE, newpage = FALSE)
popViewport()
pushViewport(viewport(layout.pos.col = 2, layout.pos.row = 2))
mosaic(art, spacing = spacing_highlighting, keep = TRUE, newpage = FALSE)
popViewport(2)
@
\caption{Varying spacing for the Arthritis  data. Top left: equal
  spacing for all dimensions, top right: different spacings for
  individial dimensions, bottom left: increasing spacing, bottom right: 
  spacing used for highlighting.}
\label{fig:spacing}
\end{center}
\end{figure}
\setkeys{Gin}{width=0.7\textwidth}
